Measuring Hash Trustworthiness via Collision Utility Metrics: Logical Cryptanalysis of MD4

### 2017-12-05 dsn-2018/bare_conf.tex
Minor changes
Signed-off-by: Alexander Scheel <alexander.m.scheel@gmail.com>

Cryptographic hash functions form the core of many protocols which
trust that cryptographic hash functions preserve certain basic
properties. Cryptograpic hash functions see widespread
use for file integrity checks to verify long term storage of
data, as cache invalidation techniques, and as a building block in network
protocols such as Kerberos and TLS.  The classes of functions suitable
for these use cases necessarily have
strong guarantees about properties of its members.
An ideal cryptographic hash has five main features: determinism - the
same message always results in the same hash; performability -
computing a hash for a message should not be algorithmically
difficult; non-reversability - it should be infeasible to generate a message
from its hash except by trying all possible messages; sensitivity to
initial conditions - small changes to a message should change the hash
value to a degree that the new hash value appears uncorrelated with
the old hash value; unique digests - it should be infeasible to find
two different messages with the same hash value.

@@

We seek to address the problem of evaluating the use cases for a collision,
to determine whether or not a collision affects a particular system.
The central thesis of this research is that logical cryptanalysis
allows us to formalize the structure inherent to the underlying hash
function such that new collisions can be found within reasonable time
on commodity hardware.  We assert that when such structure exists, it
is \textbf{imperitive to discontinue use of such hash functions as
  they are fundamentally untrustworthy for critical applications.}  An
example of such structure showing the unit distance neighborhood
(defined in Section \ref{Sec:Terminology}) of the known collisions
from Dobbertin and Kasselman is presented in Figure \ref{Fig:Graph-Intro}.
This paper contributes several techniques to logical cryptanalysis and uses these
techniques to derive metrics of the trustworthiness for hash functions. We claim
the following results are novel in the field:
\begin{itemize}
    \item We define the neighborhood of a class of collisions and show that
        it is frequently non-empty.
    \item We discuss techniques for finding useful collisions within a
        class using logical cryptanalysis.
    \item We propose additional techniques for building new classes of
        collisions and show that classes of collisions in MD4 can be found inductively.
    \item We propose five universal metrics for evaluating the utility of
        a class of collisions.
\end{itemize}
We propose that collision classes of high utility are closer to a second
preimage attack in the amount of flexibility they provide to an attacker,
whereas classes of low utility may not affect all systems which use a
particular hash function.

The remainder of this paper is organized as follows. In Section
\ref{Sec:Related} we discuss previous results and show how our research
expands upon prior work. In Section \ref{Sec:Terminology}, we discuss the
terminology and notations we use throughout the remainder of the paper.
In Section \ref{Sec:Intuition}, we give the intuition behind measuring
the utility of a collision and motivate why it is useful. In Section
\ref{Sec:Logical}, we present new techniques in logical cryptanalysis to
analyze collisions in hash functions. In Section \ref{Sec:Empirical}, we
discuss the utility of existing collisions under the metrics described in
\ref{Sec:Intuition}. In Section \ref{Sec:Collision}, we discuss new
differential paths and the effectiveness of various techniques presented
in Section \ref{Sec:Logical}. Lastly, in Section \ref{Sec:Impacts}, we
discuss the broader impacts and conclusions of this research, and
conclude with Section \ref{Sec:Conclusion} discussing future work and
extensions in Section \ref{Sec:Future}.

@@

Table \ref{table:differentials} illustrates the number of unique differentials,
including the input block differential specified by the original authors, which produces a differential path conforming to
each of the collision classes. We note in particular that while Y. Sasaki's
attack is generally an improvement upon X. Wang's, Y. Sasaki's attack has fewer
input block differentials and thus less flexibility in this aspect. Further,
M. Schl{\"a}ffer's attack -- also building off the work of X. Wang -- has the
same number of differentials and is thus also more flexible than Y. Sasaki's.

@@

Table \ref{table:neighborhood} presents the sizes discovered empirically of neighborhoods
of existing collisions. In particular, note that both Y. Sasaki's attack and
M. Schl{\"a}ffer's attack -- which both claimed to improve upon X. Wang's
attack--have larger neighborhood sizes, and likewise with P. Kasselman's
attack which improved upon H. Dobbertin's attack. Thus, the newer collisions
occur in more dense subsets of the collision class space.

@@

This last point shows that, while there may be a number of unique differentials
for a given collision class, and potentially many collision classes within
some neighborhood of the original collision class, by analyzing the structure
of the input block differential, systems can potentially prohibit successful
collision attacks by preventing structures of input block differences. Note
too how the graph presented in Figure \ref{Fig:Families48} shows a distinct grouping
into three disconnected subgraphs, which fall along input block differential
family boundaries.

@@

We have demonstrated in this paper that our theories for logical
cryptanalysis hold, and lead to a 35 fold increase in collision
discovery when compared to the existing literature.  By employing our
presented distance heuristic and the lattice structure uncovered by
the $\precsim$ relation, we can evaluate the trustworthiness of a hash
function as the inverse of the flexibility of known collisions for
that hash function.  This allows us to measure the structural
weaknesses of a hash function and determine if extant collisions are sufficient
to render a hash function untrustworthy, and likewise to understand if
necessary conditions related to the underlying structure are met to
ensure trustworthiness.

We demonstrate the utility of five metrics of the flexibility present
in a collision on the basis of 1) the number of unique differentials
within a collision class; 2) the size of the unit distance
neighborhood for a collision class; 3) the maximum count of zeros in a
binary representation of a colliding block; 4) the existance of blocks
which collide under multiple initial values; and 5) the presence of
ASCII valued inplut blocks within a collision.  These metrics proved
capable when evaluating the flexibility of collisions in the published
literature, and showed that flexibility is a multi-dimensional
quantity that must be compared over multiple distance metrics, as not all
collisions dominate universally.

Our empirical results furthermore demonstrated that concerns over
industrial use of deprecated hash functions, specifically MD4, are
real, and that structural weaknesses exist which are highly
exploitable and limit the trustworthiness of any system which employs
such hash functions.  We hope the presented results lead to a
reduction in the use of such deprecated hash functions, and that new
hash functions are periodically evaluated when collisions are
discovered to understand if these collisions also indicate the
presence of structural weaknesses.

>>>>

Cryptographic hash functions form the core of many protocols which
trust that these functions preserve certain basic
properties. Cryptograpic hash functions see widespread
use as file integrity checks to verify long term storage of
data, as cache invalidation techniques, and as a building block in network
protocols such as Kerberos and TLS.  The classes of functions suitable
for these use cases necessarily have
strong guarantees about properties of its members.
An ideal cryptographic hash has five main features: determinism - the
same message always results in the same hash; performability -
computing a hash for a message should not be algorithmically
difficult; non-reversability - it should be infeasible to generate a message
from its hash except by trying all possible messages; sensitivity to
initial conditions - small changes to a message should change the hash
value to a degree that the new hash value appears uncorrelated with
the old hash value; unique digests - it should be infeasible to find
two different messages with the same hash value.

@@

We seek to address the problem of evaluating the use cases for a collision,
to determine whether or not a collision affects a particular system.
The central thesis of this research is that logical cryptanalysis
allows us to formalize the structure inherent to the underlying hash
function such that new collisions can be found within reasonable time
on commodity hardware.  We assert that when such structure exists, it
is \textbf{imperitive to discontinue use of such hash functions as
  they are fundamentally untrustworthy for critical applications.}  An
example of such structure showing the unit distance neighborhood
(defined in Section \ref{Sec:Terminology}) of the known collisions
from Dobbertin and Kasselman is presented in Figure \ref{Fig:Graph-Intro}.

This paper contributes several techniques to logical cryptanalysis and uses these
techniques to derive metrics of the trustworthiness for hash functions. We claim
the following results are novel in the field:
\begin{itemize}
    \item We define the neighborhood of a class of collisions and show that
        it is frequently non-empty.
    \item We discuss techniques for finding useful collisions within a
        class using logical cryptanalysis.
    \item We propose additional techniques for building new classes of
        collisions and show that classes of collisions in MD4 can be found inductively.
    \item We propose five universal metrics for evaluating the utility of
        a class of collisions.
\end{itemize}
We propose that collision classes of high utility are closer to a second
preimage attack in the amount of flexibility they provide to an attacker,
whereas classes of low utility may not affect all systems which use a
particular hash function.

The remainder of this paper is organized as follows. In Section
\ref{Sec:Related} we discuss previous results and show how our research
expands upon prior work. In Section \ref{Sec:Terminology}, we discuss the
terminology and notations we use throughout the remainder of the paper.
In Section \ref{Sec:Intuition}, we give the intuition behind measuring
the utility of a collision and motivate why it is useful. In Section
\ref{Sec:Logical}, we present new techniques in logical cryptanalysis to
analyze collisions in hash functions. In Section \ref{Sec:Empirical}, we
discuss the utility of existing collisions under the metrics described in
\ref{Sec:Intuition}. In Section \ref{Sec:Collision}, we discuss new
differential paths and the effectiveness of various techniques presented
in Section \ref{Sec:Logical}. Lastly, we discuss the broader impacts and
conclusions of this research in Section \ref{Sec:Conclusion}, discussing future work and
extensions in Section \ref{Sec:Future}.

@@

Table \ref{table:differentials} illustrates the number of unique differentials,
including the input block differential specified by the original authors, which produces a differential path conforming to
each of the collision classes. We note in particular that while Y. Sasaki's
attack is generally an improvement upon X. Wang's, Y. Sasaki's attack has fewer
input block differentials and thus less flexibility in this aspect. Further,
M. Schl{\"a}ffer's attack --- also building off the work of X. Wang --- has the
same number of differentials and is thus also more flexible than Y. Sasaki's.

@@

Table \ref{table:neighborhood} presents the sizes discovered empirically of neighborhoods
of existing collisions. In particular, note that both Y. Sasaki's attack and
M. Schl{\"a}ffer's attack --- which both claimed to improve upon X. Wang's
attack---have larger neighborhood sizes, and likewise with P. Kasselman's
attack which improved upon H. Dobbertin's attack. Thus, the newer collisions
occur in more dense subsets of the collision class space.

@@

This last point shows that, while there may be a number of unique differentials
for a given collision class, and potentially many collision classes within
some neighborhood of the original collision class, by analyzing the structure
of the input block differential, systems can potentially prohibit successful
collision attacks by preventing structures of input block differences. Note
too how the graph presented in \ref{Fig:Families48} shows a distinct grouping
into three disconnected subgraphs, which fall along input block differential
family boundaries.

@@

- Theory behind Logical Cryptanalysis:
    - Distance heuristic.
    - $\precsim$.
- Tie into Trustworthiness of hash functions.
- Demonstrated metrics.
- Demonstrated 35k new collisions.
- Implications to real world systems.
